---
title: SCHOOL-DISTRICT-DATA-TRACY
description: |
  Data Analysis of Massachusetts Public Schools' Class Size by Gender and Selected Populations for DACSS601 Final Project, HW4 and HW5
Author: Erin-Tracy
date: 08-24-2021
output:
  distill::distill_article:
    self_contained: no
draft: yes
---

**INTRODUCTION**

"The Class Size by Gender and Selected Populations data represents the number of classes and average class size for each subject by gender, Limited English Proficient and Low Income"^1 for public school districts in Massachusetts.

According to this data, Massachusetts Public School Districts have on average 1142 total number of classes with an average class size of 17.7. The average Massachusetts Public School District has 2382 students with 48.8% identifying as Female and 50.9% identifying as Male. On average, 6.66% of students are English Language Learners (ELLs), 18.47% of students have disabilities and 30.53% are Economically Disadvantaged.

The purpose of this analysis is to compare the data from school districts with a high number of economically disadvantaged students to districts with a low number of economically disadvantaged students. My prediction is that schools with larger percentages of economically disadvantaged students will also have larger populations of students with disabilities and of English Language Learners, as well as larger class sizes. I predict that percentages of male and female students will be the same regardless of the economically disadvantaged variable.

I think this is important data to study because if my predictions are correct, further research should be done to look into the breakdown of state funding for districts with greater numbers of economically disadvantaged students. I am assuming that districts with more economically disadvantaged students do not receive the same amount of local funding per student as districts with fewer economically disadvantaged students, and so they would rely on the state to even the playing field. If school districts with greater numbers of economically disadvantaged students do in fact also have more ELLs and students with disabilities, those districts actually need more funding per student, given that students who are ELLs and students with disabilities need more support services than other students.

If my predictions are correct that class sizes are larger among districts with more economically disadvantaged students, again all signs point to more research on funding. Class sizes can only decrease with more staff and more space, both of which require additional funding. For more information on the benefits of smaller class size and student learning, please refer to the multiple sources available at the website below.

https://classsizematters.org/research-and-links/  ^2

I am the daughter of a public school teacher, the mother of a public school Pre-Kindergartener and an employee of the state's largest teacher's union. I am passionately pro-public education. I think we all need to constantly be checking to make sure that public school students across the state are all receiving a quality education. Comparing the variables of this data set is one way to see if students are having different experiences, and if those differences are at all related to economic situation.


**DATA**

The 9 variables are District, Total Number of Classes (Classes), Average Class Size (avgclasssize), Number of Students (students), Percentage of Students who identify as Male (malepercent), Percentage of Students who identify as Female (femalepercent), Percentage of students that are English Language Learners (ellpercent), Percentage of Students with Disabilities (studentswdisabilitiespercent) and Percentage of students that are Economically Disadvantaged (econdisadvantagedpercent). Number of students includes Pre-Kindergarten through 12th grade students. All "Percentage of" student data is out of the entire school district population. Additional definitions are included below.

There are 403 observations (school districts). Please note that only public school districts are included. This includes some public charter schools, but does not include private charter schools or any private schools in Massachusetts. Some observations include missing data.

More information is also available at the original source: https://profiles.doe.mass.edu/statereport/classsizebygenderpopulation.aspx

Data sourced from SIMS (Student Information Management Systems), SCS (Student Course Schedule) and EPIMS (Education Personnel Information Management System)

**DEFINITIONS**

"Total # of Classes: Number of classes is based on classes that could be linked between SIMS, EPIMS, and SCS by School, Course, Section, and Term. The class must have a class count greater than 1 (one) and have students in the class who have a SCS enrollment status of enrolled, completed, or incomplete."^1

"Average Class Size: Average Class Size is calculated by dividing the total number of students in classes by the total number of classes. Students taking multiple classes will be included in multiple class size averages."^1

"Number of Students: This figure is de-duplicated count of students at the subject, district, and school levels."^1

"English Learners: Indicates the percent of enrollment who are English learners, defined as "a student whose first language is a language other than English who is unable to perform ordinary classroom work in English."^1

"Economically Disadvantaged: Calculated based on a student's participation in one or more of the following state-administered programs: the Supplemental Nutrition Assistance Program (SNAP); the Transitional Assistance for Families with Dependent Children (TAFDC); the Department of Children and Families' (DCF) foster care program; and MassHealth (Medicaid)."^1

The Massachusetts Department of Elementary and Secondary Education (DESE) is the state education agency for Massachusetts. One of DESE's primary purposes is to collect data to inform state and local decisions. MA DESE's website includes hundreds of different data sets. ^3

**PACKAGES**

In order to explore the data, I installed the following packages.


```{r setup, include=TRUE}
knitr::opts_chunk$set(echo = TRUE)

library("tidyverse")
library("readr")
library("ggplot2")
library("dplyr")

```


**READING IN, CLEANING DATA & DEALING WITH MISSING DATA: HOMEWORK 4**

While reading in this data, I did remove the School District ID column, which may be useful in future research. (Perhaps if someone was interested in combining various DESE datasets into one.)

I also had some observations that included "0" as a value for all variables.That seemed to be meaningless since a district could not have 0 students, 0 classes, etc. I was concerned that keeping that data in the dataset would incorrectly skew my analysis. Including a "0" in a small data set, when calculating average would result in an incorrect low average. Thanks to the #r-help slack page, I was able to receive advice as to what code would help me correct this issue of missing data (that was not initially labelled as NA).

I did include a preview of this missing data below, to get a sense of how many districts are missing information. Gosnold was the only district that included 0 for all variables.


```{r}
data<-read_csv("ClassSizebyGenPopulation.csv",skip=2,
               col_names= c("District", "delete", "classes", "avgclasssize", "students","femalepercent","malepercent","ellpercent","studentswdisabilitiespercent","econdisadvantagedpercent"))%>%
  select(!contains("delete"))

data.missing <- data%>%
  mutate(allna = rowSums(select(., classes:econdisadvantagedpercent)),
         across(classes:econdisadvantagedpercent, ~case_when(
           . == 0 ~ NA_real_,
           . > 0 ~ .)))%>%
  filter(allna ==0)%>%
  select(-allna)

head(data.missing)

data.clean <- data%>%
  mutate(allna = rowSums(select(., classes:econdisadvantagedpercent)),
         across(classes:econdisadvantagedpercent, ~case_when(
           . == 0 ~ NA_real_,
           . > 0 ~ .)))%>%
  filter(!allna ==0)%>%
  select(-allna)

head(data.clean)

```
Here I ran some very basic code to help me understand this data.

```{r}
summarise_all(data.clean, mean)

dim(data.clean)

colnames(data.clean)

head(data.clean)

tail(data.clean)

```

**VISUALIZATION: HOMEWORK 5**

I created 3 graphs using ggplot to provide preliminary visual graphics to help me answer the specific question, "are class sizes larger among school districts with the greatest number of economically disadvantaged students?" The geomsmooth ggplot graph suggests "yes, slightly", but the ggplots with geom_points are less decisive. Clearly further analysis is necessary.


```{r}
ggplot(data = data.clean)+
  geom_point(mapping = aes(x= econdisadvantagedpercent, y=avgclasssize))+
  labs(title = "MA Public School Data Graph 1, Geom Point" , x = "Percentage of Economically Disadvantaged Students", y= "Average Class Size")


ggplot(data = data.clean)+
  geom_smooth(mapping = aes(x= econdisadvantagedpercent, y=avgclasssize))+
  labs(title = "MA Public School Data Graph 2, Geom Smooth" , x = "Percentage of Economically Disadvantaged Students", y= "Average Class Size")

ggplot(data = data.clean) +
  geom_point(mapping = aes(x = econdisadvantagedpercent, y = avgclasssize)) +
  geom_smooth(mapping = aes(x = econdisadvantagedpercent, y = avgclasssize))+
  labs(title = "MA Public School Data Graph 3, Geom Point & Smooth" , x = "Percentage of Economically Disadvantaged Students", y= "Average Class Size")

```


I found the ggplot with geompoint and geomsmooth to provide the most useful visual for this type of question. I repeated the same type of ggplot graph for the question "are there more students with disabilities at schools with the greatest number of economically disadvantaged students?" Again, it looks like the answer is yes, but only slightly.

```{r}

ggplot(data = data.clean) +
  geom_point(mapping = aes(x = econdisadvantagedpercent, y = studentswdisabilitiespercent)) +
  geom_smooth(mapping = aes(x = econdisadvantagedpercent, y = studentswdisabilitiespercent))+
  labs(title = "MA Public School Data Graph 4, Geom Point & Smooth" , x= "% of Economically Disadvantaged Students", y = "% w Disabilities")

```

The final ggplot graph in this section addresses the question, "are there more Engligh Language Learners at schools with the greatest number of economically disadvantaged students?" The answer to this question seems to be a more certain "yes". The Percentage of ELL students definitely seems to increase as the percentage of economically disadvantaged students increases.


```{r}
ggplot(data = data.clean) +
  geom_point(mapping = aes(x = econdisadvantagedpercent, y = ellpercent)) +
  geom_smooth(mapping = aes(x = econdisadvantagedpercent, y = ellpercent))+
  labs(title = "MA Public School Data Graph 5, Geom Point &  Smooth " , x= "% of Economically Disadvantaged Students", y = "% of ELLs")

```

**Data Exploration: Low % of Economically Disadvantaged Students vs High % of Economically Disadvantaged Students**

I created a subset of data for the 40 districts (roughly 10%) with the most economically disadvantaged students (high_econdis_data) and another for the 40 districts (roughly 10%) with the least economically disadvantaged students (low_econdis_data). I ran some quick averages to see some comparisons.

The "Low Economically Disadvantaged" district data has an average of 5.70% economically disadvantaged students, 2.22% English Language Learners and 15.1% students with disabilities.

Comparatively, the "High Economically Disadvantaged" district data has an average of 69.5% economically disadvantaged students, 19.7% English Language Learners and 22% students with disabilities.

This all seems consistent with the ggplots above, just more precise.

The "Low Economically Disadvantaged" districts have an average class size of 17.6 and 2268 total students in the district. The "High Economically Disadvantaged" districts have an average class size of 19.4 and 3852 total students in the district.

One thing I found interesting, and unexpected, was regarding gender identification.

Among the "High Economically Disadvantaged" districts, 49.6% identify as female and 50.3% identify as male, leaving approximately 0.1% of students as not identifying strictly as male or female. In the "Low Economically Disadvantaged" districts 48.4% identify as female and 49.0% identify as male, leaving approximately 2.6% as not identifying strictly as male or female.

I initially only included 10 districts in each of these subsets, but it seemed like such a small subset would not be as useful as each subset would be made up primarily of outliers.

```{r}
data<- arrange(data.clean,desc(`econdisadvantagedpercent`))

high_econdis_data<- head(data.clean,40)
low_econdis_data<- tail(data.clean,40)

summarise_all(high_econdis_data, mean)
summarise_all(low_econdis_data, mean)

low_econdis_data%>%
  select(ellpercent, studentswdisabilitiespercent, econdisadvantagedpercent)%>%
  summarize_all(mean)

high_econdis_data%>%
  select(ellpercent, studentswdisabilitiespercent, econdisadvantagedpercent)%>%
  summarize_all(mean)

low_econdis_data%>%
  select(avgclasssize, students)%>%
  summarize_all(mean)

high_econdis_data%>%
  select(avgclasssize, students)%>%
  summarize_all(mean)

low_econdis_data%>%
  select(femalepercent, malepercent)%>%
  summarize_all(mean)

high_econdis_data%>%
  select(femalepercent, malepercent)%>%
  summarize_all(mean)

```

**Visualization Cont.**

I felt limited with regard to what graphics I could produce because all of the values were integers.
I chose to create a different set of categories of districts based on percentage of Economically Disadvantaged student percentage.

VeryHigh= 60%+ Economically Disadvantaged Students

High= 40%-59.9% Economically Disadvantaged Students

MidRange= 20%-39.9% Economically Disadvantaged Students

Low= 10-19.9% Economically Disadvantaged Students

VeryLow= under 10% Economically Disadvantaged Students

These groupings were determined without a scientific process. They just made sense to me. I look forward to learning more about how it would best make sense to split values up, similar to this.

I created a table and a prop table to see the how the numbers and percentages brokedown among these different categories. The MidRange group definitely has the majority (36%). The groups get progressively smaller as you move towards the extreme values.

Initially the categories were not in a logical order, but in alpha order (High, MidRange, Low...) Thank you to Sean for helping me reorder.

```{r}
data<-data.clean%>%
  mutate(EconDisadvantagedGroup = case_when(
         econdisadvantagedpercent >= 60 ~ "VeryHigh",
         econdisadvantagedpercent >= 40 & econdisadvantagedpercent < 60 ~ "High",
         econdisadvantagedpercent >= 20 & econdisadvantagedpercent < 40 ~ "MidRange",
         econdisadvantagedpercent >= 10 & econdisadvantagedpercent < 20 ~ "Low",
         econdisadvantagedpercent < 10 ~ "VeryLow")) %>% mutate(EconDisadvantagedGroup=factor(EconDisadvantagedGroup, levels=c("VeryHigh", "High", "MidRange", "Low", "VeryLow")))


head(data)
colnames(data)

table(data$EconDisadvantagedGroup)

data%>%
  select(EconDisadvantagedGroup)%>%
  table()%>%
  prop.table()*100

```
The work I did above was necessary set up so that I could to create a ggplot graph with facet wraps.

I think these visuals are very useful. They show that districts with the smallest amounts of economically disadvantaged students (Low and Very Low) have very small amounts of English Language Learners and consistently have small to average class sizes.

On the contrary, the schools with high percentages of economically disadvantaged students (High and Very High) have much higher percentages of English Language Learners and a wider range of class sizes. Fortunately we do not see the districts with the highest numbers of English Language Learners also having the largest class sizes.

```{r}
ggplot(data=data)+
  geom_point(mapping=aes(x= ellpercent, y= avgclasssize))+
  facet_wrap(~EconDisadvantagedGroup,nrow=2)+
  scale_y_continuous(limits=c(10,45), breaks=seq(0,50,5))+
  labs(title = "MA Public School divided by Economically Disadvantaged Group" , x= "% of English Language Learners", y = "AVerage Class Size")

```

**REFLECTION**

It was rewarding to work with data I am interested in. I was pleased that I discovered some things that I didn't expect to (high percentage of students with neither male nor female gender identification at schools with fewer economically disadvantaged students, and districts with highest levels of ELL students do not also have the largest class sizes.)  

There are some things I would have done differently if I were to start this project over now. I should have given the data a more unique name (than data). I need to put more thought into the names I assign to columns. Some of the names I used were very long, I was concerned that if I shortened them, the reader wouldn't know what data I was referencing. I also need to do a better job using punctuation or capital letters to separate words in the variable names. In the code, I have not applied spaces consistently. I hadn't gone back and made all of these changes to my current data, because I worried that I would inadvertently create errors.

I'm worried that my inexperienced attempt to categorize the data (specifically Economically Disadvantaged student percentages) might create different results than if I had a better understanding of how to do this step. I chose my own data set. Next time I will probably select a data set from one of the sources mentioned in class. I would love to start with a data set that has more types of variables than just integers and percentages (like with the diamond data, variables like cut, quality...) I felt limited with regard to which graphs I could use. I would love to experiment with the Fill feature in ggplot.

It was surprising which parts of this project were time consuming. Things that I thought would be quick and easy, took a long time. And some things that I thought would be time consuming, ended up being quick and easy. I did learn that I really need to understand every piece of code in my project. We spent a significant amount of time trying to figure out why my facet wraps were extremely small. (Thank you Meredith and Sean for finding the unnecessary code causing that problem just under my YAML.)

8-12 pages seemed like it would be very long, but I find myself with approximately 16 pages. It will take me some time to figure out how much code and explanation corresponds to 1 page, etc.

As mentioned above, I wish I could have more scientifically created categories for variables with numerical values. I wish I had a better understanding of the Chi Test process to be able to test these findings, to see if they are significant.

I may have tried to include too many variables for my first project, but if I were to continue with this project, I would have like to incorporate another set of data for the same school districts that includes state and local funding. It would be very interesting to see how much money the school districts receive for each student on average, and compare that among the wealthy and not wealthy districts.

I would also love to compare this current data to the same data from 20-30 years ago. What is the same, and what has changed (significantly)? Have some districts seen a greater increase in population than others? And what variables have changed? Are there significantly more ELL students? Or more economically disadvantaged students? Have the standards for determining those values also changed over time?


**Conclusion**

Is the student experience different in school districts with greater numbers of economically disadvantaged students?

The answer is yes. I compared the 40 districts with the most economically disadvantaged students with the 40 districts with the least economically disadvantaged students. (The former had more than 10 times as many disadvantaged students as the latter.) For ease, we'll refer to them as the wealthiest districts and the least wealthy districts. The least wealthy districts have 8 times as many English Language Learners as the wealthy districts. This was the most significant difference. The least wealthy districts have 50% more students with disabilities than the wealthy districts. The average class sizes of the least wealthy districts are 10% larger than the wealthy districts.

Specifically regarding the correlation between percentage of English Language Learners and Economically Disadvantaged Students,I revisited the entire data set to make sure that focusing in on 20% of the data set wasn't providing me with false information. After assigning a category to all observations based on economically disadvantaged status, I studied a group of facet wraps. They confirmed my initial finding that there are significantly more English Language Learners in schools with more Economically Disadvantaged students. This information would not be shocking to anyone, I'm sure that in general, in the United States people who are not proficient in English have lower incomes than those that are, so children who are not proficient in English are also more likely to be Economically Disadvantaged.

I hope these findings would encourage us to look more deeply into how much funding and other resources are extended specifically to the schools with high rates of English Language Learners. Every attempt should be made to give all Public School Students in Massachusetts the tools they need to graduate from high school and have successful careers. If we don't we will continue to see a larger achievement gap and more income inequality.


**Bibliography**

[1: https://profiles.doe.mass.edu/statereport/classsizebygenderpopulation.aspx]

[2: https://classsizematters.org/research-and-links/]

[3: https://www.doe.mass.edu/About.html]


Distill is a publication format for scientific and technical writing, native to the web.

Learn more about using Distill at <https://rstudio.github.io/distill>.
