---
title: "manninghwk4"
description: |
  Univariate Analysis
author: Michelle Manning
date: 08-28-2021
output:
  distill::distill_article:
    self_contained: false
draft: true
---

I am looking at the data set "Crime in Los Angeles Data from 2020 to Present". I have heard that 2020 had an interesting rise in certain crimes like murder, but a decrease in crime like bunco. The pandemic surely had a hand in this, but I wondered what 2021 crime data looked as covid restrictions were lifted in comparison to the same 2020 period.

Questions I intend to explore: 
1. How many crimes occurred in the first seven months of 2020 vs 2021?
2. What is the type of crime that occurred in 2020 vs 2021?
3. Which areas are being affected?  
4. Out of the community area with highest crime increase in 2021, what types of crime occurred?
           
First I changed echo to TRUE and loaded in the libraries tidyverse and stringr. Next I read in the csv file from Kaggle.com and viewed it to examine the data.
```{r setup, echo=TRUE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(stringr)
LAcrime <- read_csv("Crime_Data_from_2020_to_Present.csv")
```

Cleaning the Data set:
First bit of cleaning necessary was to get rid of all the variables that I wouldn't be looking at and renamed the variables left. Next, I checked for any NAs or nulls. I removed no nulls, but a few NAs. Next, I separated the date variable into day, month, and year variables while also removing the incorrect time attached to the date variable. I wanted to look at 2020 and 2021 in an equal time period, so I decided to focus on the first seven months of each year to answer my questions.
```{r, echo=TRUE}
LAcrime2 <- LAcrime %>% 
  select(`DATE OCC`, AREA, `AREA NAME`, `Crm Cd`, `Crm Cd Desc`) %>% 
  rename(DateOcc=`DATE OCC`, Area=AREA, AreaName=`AREA NAME`, CrmCd=`Crm Cd`, CrmDesc=`Crm Cd Desc`)#Selecting columns needed & renaming them without spaces in column name.

any(is.na(LAcrime2))
any(is.null(LAcrime2))#Checking for NAs & nulls
map(LAcrime2, ~sum(is.na(.)))
LAcrime3 <- na.exclude(LAcrime2)#Get rid of NAs and nulls
#Breaking up the dates into separate columns to better order the data and focus on the observations with regard to the year
LAcrime4 <- LAcrime3 %>% 
  separate(DateOcc, sep="/", into = c("OCCmonth", "OCCday", "OCCyear")) %>% 
  separate(OCCyear, sep=" ", into = c("OCCyear", "1","2")) %>% 
  select(-"1", -"2")
LAcrime5 <- filter(LAcrime4,!(OCCmonth>= "08"))#Creating the same time period window for 2020 as 2021. 7 months, Jan 1st-July 31st.
```
Reshaping the variables to fit my scope:
The amount of crime types is unweildy, so I am focusing on the top 10 crimes in 2020 and 2021. Next, I shaped the data to reflect the amount of crimes in each community area. Lastly, I did the same for the specific community area: Pacific. I will also only focus on the top 10 crimes in the Pacific area because the graph of all crime types is barely legible. 
```{r}
# Top 10 crimes 
n_distinct(LAcrime5$CrmDesc) #Check
crmdesc2020_1 <- LAcrime5 %>%
   filter(OCCyear=="2020") %>% 
  count(CrmDesc) %>% 
  arrange(desc(`n`)) %>% 
  rename(count=`n`) #Count of crimes in 2020 added to this data set
Top10_2020 <- crmdesc2020_1[1:10, ] #This object is the Top 10 crime types for 2020
crmdesc2021_1 <- LAcrime5 %>%
   filter(OCCyear=="2021") %>% 
  count(CrmDesc)%>% 
  arrange(desc(`n`))%>% 
  rename(count=`n`)#Count of crimes in 2021 added to this data set
Top10_2021 <- crmdesc2021_1[1:10, ]#This object is the Top 10 crime types for 2021
#Putting a column of the years back on to data set correctly
year2020 <- c(rep(2020,10))
year2021 <- c(rep(2021,10))
LAcrimeTop10_1 <- cbind(Top10_2020, year = year2020) 
LAcrimeTop10_2 <- cbind(Top10_2021, year = year2021) 
#Combining these data sets into one Top 10 across both years.
LAcrimeTop10_3 <- rbind(LAcrimeTop10_1, LAcrimeTop10_2)
LAcrimeTop10_3$year <- as.factor(LAcrimeTop10_3$year)  #Changing the new year column into a factor, so it isn't read as a continuous variable for the plots later

#Community Areas
n_distinct(LAcrime5$AreaName)#Check
Area2020 <- LAcrime5 %>%
   filter(OCCyear=="2020") %>% 
  count(AreaName) %>% 
  arrange(desc(`n`))%>% 
  rename(count=`n`)#Count of Area crimes in 2020 added to this data set
Area2021 <- LAcrime5 %>%
   filter(OCCyear=="2021") %>% 
  count(AreaName)%>% 
  arrange(desc(`n`)) %>% 
  rename(count=`n`)#Count of Area crimes in 2021 added to this data set
yearArea20 <- c(rep(2020,21))
yearArea21 <- c(rep(2021,21))#Putting a column of the years back on to data set correctly
LAcrime_Area1 <- cbind(Area2020, year = yearArea20) 
LAcrime_Area2 <- cbind(Area2021, year = yearArea21) 
#Combining these data sets into one Crime Area set across both years.
LAcrime_Area3 <- rbind(LAcrime_Area1, LAcrime_Area2)
LAcrime_Area4 <- LAcrime_Area3 %>% 
  arrange(AreaName)#Arranged to show the most crimes vs the area they happened in
LAcrime_Area4$year <- as.factor(LAcrime_Area4$year) #Changing the new year column into a factor, so it isn't read as a continuous variable for the plots later

#Crime Types in the Pacific Area
Pacific2020 <- LAcrime5 %>%
  select(OCCyear, AreaName, CrmDesc) %>% 
   filter(OCCyear=="2020" & AreaName== "Pacific") %>% 
  count(CrmDesc) %>% 
  arrange(desc(`n`))%>% 
  rename(count=`n`)#Count of crimes in the Pacific area in 2020 added to this data set
Pacific2020_1 <- Pacific2020[1:10, ]#This object is the Pacific crime types for 2020
Pacific2021 <- LAcrime5 %>%
  select(OCCyear, AreaName, CrmDesc) %>% 
   filter(OCCyear=="2021" & AreaName== "Pacific") %>% 
  count(CrmDesc) %>% 
  arrange(desc(`n`))%>% 
  rename(count=`n`)#Count of crimes in the Pacific area in 2021 added to this data set
Pacific2021_1 <- Pacific2021[1:10, ]#This object is the Pacific crime types for 2021
#Putting a column of the years back on to data set correctly
yearPacific20 <- c(rep(2020,10))
yearPacific21 <- c(rep(2021,10))
LAcrimePacific1 <- cbind(Pacific2020_1, year = yearPacific20) 
LAcrimePacific2 <- cbind(Pacific2021_1, year = yearPacific21)
#Combining these data sets into one Pacific crimes set across both years.
LAcrimePacific3 <- rbind(LAcrimePacific1, LAcrimePacific2)
LAcrimePacific3$year <- as.factor(LAcrimePacific3$year)#Changing the new year column into a factor, so it isn't read as a continuous variable for the plots later
```

Plots:
```{r, echo=TRUE}
#Plot that shows the crime count difference between 2020 and 2021
ggplot(LAcrime5, aes(x=OCCyear))+ 
  geom_bar()
#plot that shows the top 10 crimes across both 2020 & 2021
ggplot(LAcrimeTop10_3, aes(x= year,y=count, fill= CrmDesc))+ 
  geom_bar(position="dodge", stat="identity")+
  geom_text(aes(label = CrmDesc), size = 3, hjust = 1.05, position = position_dodge(.9),angle = 90, colour = "white")+
  labs(title="Number of Crimes Committed by Type in 2020 & 2021",
        x ="Year", y = "Crime Occurences", fill ="Crime Type")
#Plot that shows the crime count by Area across both 2020 & 2021
ggplot(LAcrime_Area4, aes(x=AreaName, y=count, fill= year))+
  geom_bar(position="dodge", stat="identity")+
  geom_text(aes(label = count),size=3, vjust = 1.5,position = position_dodge(.95), colour = "white")+
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))+
  labs(title="Number of Crimes Committed by Community Areas in 2020 & 2021",
        x ="Community Area", y = "Crime Occurences", fill ="Year")
#plot that shows the crime count in the Pacific area specifically
ggplot(LAcrimePacific3, aes(x= CrmDesc,y=count, fill = year))+
  geom_bar(position="dodge", stat="identity")+
  theme(axis.text.x = element_text(angle = 90,size=10, vjust = 0.5, hjust=1))+
  labs(title="Types of Crimes Committed in the Pacific Area in 2020 & 2021",
        x ="Crime Type", y = "Crime Occurences", fill ="Year")
```

