---
title: "homework4_zutima"
description: |
  A short description of the post.
author:
  - name: Abhinav Kumar
    url: {}
date: 08-25-2021
output:
  distill::distill_article:
    self_contained: false
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```
```{r}
library(tidyverse)
library(stringr)
library(dplyr)

```
Tasks for Assignment 4

For this assignment, you must select at least two variables of interest from your dataset, provide a basic description of the variables, clean and recode as needed, and present descriptive statistics and any visualizations. Your R Markdown document should include:

1. Descriptions of the variables - how they were collected, any missing values, etc
2. How you cleaned and coded the data, including a before/after comparison as needed
3. summary descriptives of the recoded variables
4. appropriate visualizations (not required)

```{r Read in the data}

book <-read_csv(file="../../_data/hotel_bookings.csv")

```

### Now we will read in the data. For this assignment, I am looking at the hotel bookings data which shows us data describing the best time and year to book a hotel. Source for this data is kaggle: https://www.kaggle.com/jessemostipak/hotel-booking-demand


```



