---
title: "AnkittProjectDacss601"
author: "Ankit"
date: "8/23/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Loading relevant packages

```{r}
library(tidyverse)
```

Reading the dataset into a data frame

```{r}
library(readr)
life_expectancy <- read_csv("_data/UNdata_Export_20210825_221659882.csv")
```

Taking a look at the first few rows

```{r}
head(life_expectancy)
```

Subsetting the life expectancy data

```{r}
subdata <- life_expectancy  %>% 
    filter(Year == "2000-2005") %>%
    select("Country or Area", "Subgroup", "Value")

head(subdata)
```
Reshaping the life expectancy data

```{r}
subdata <- pivot_wider(subdata, names_from = "Subgroup", values_from = "Value")

head(subdata)
```

Plotting male and female life expectancy

```{r}
ggplot(subdata, aes(x = Male, y = Female)) +
  geom_point()
```

Adding an abline and changing the scale of axes of the previous plots
```{r}
ggplot(subdata, aes(x = Male, y = Female)) +
  geom_point() + geom_abline(slope = 1, intercept = 0, linetype = "dashed") +
xlim(35, 85) + ylim(35,85)
```

Adding labels to previous plot

```{r}
ggplot(subdata, aes(x=Male, y=Female))+
  geom_point(colour="white", fill="chartreuse3", shape=21, alpha=.55, size=5)+
  geom_abline(intercept = 0, slope = 1, linetype=2)+
  scale_x_continuous(limits=c(35,85))+
  scale_y_continuous(limits=c(35,85))+
  labs(title= "Life Expectancy at Birth by Country",
       subtitle="Years. Period: 2000-2005. Average.",
       caption= "Source: United Nations Statistics Division",
       x= "Males",
       y="Females")
```

Subseting data to obtain countries of interest
```{r}
top_male <- subdata %>% arrange(Male-Female) %>% head(3)
top_female <- subdata %>% arrange(Female-Male) %>% head(3)

top_male
top_female
```

Adding text to the previous plot to label countries of interest

```{r}
ggplot(subdata, aes(x=Male, y=Female))+
  geom_point(colour="white", fill="chartreuse3", shape=21, alpha=.55, size=5)+
  geom_abline(intercept = 0, slope = 1, linetype=2)+
  scale_x_continuous(limits=c(35,85))+
  scale_y_continuous(limits=c(35,85))+
  labs(title="Life Expectancy at Birth by Country",
       subtitle="Years. Period: 2000-2005. Average.",
       caption="Source: United Nations Statistics Division",
       x="Males",
       y="Females") +
    geom_text(data = top_male, aes(x = "Male", y = "Female", label = "Country or Area"))+
    geom_text(data = top_female, aes(x = "Male", y = "Female", label = "Country or Area"))+    
    theme_bw()
```

