---
title: "hotel_Bookings"
author: "sathvik_thogaru"
date: "11/08/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## R Markdown

# context

This data set contains booking information for a city hotel and a resort hotel, and includes information such as when the booking was made, length of stay, the number of adults, children, and/or babies, and the number of available parking spaces, among other things.

Have you ever wondered when the best time of year to book a hotel room is? Or the optimal length of stay in order to get the best daily rate? What if you wanted to predict whether or not a hotel was likely to receive a disproportionately high number of special requests?

```{r}
library(readr)
hotel_bookings <- read_csv("_data/hotel_bookings.csv")
```


```{r}
head(hotel_bookings)
tail(hotel_bookings)
str(hotel_bookings)
```


```{r}
dim(hotel_bookings)
summary(hotel_bookings)
```



```{r}
colnames(hotel_bookings)
```


```{r}
View(hotel_bookings)
```

```{r}
library(dplyr)
```

```{r}
unique(hotel_bookings$hotel)
unique(hotel_bookings$is_canceled)
unique(hotel_bookings$meal)
unique(hotel_bookings$country)
unique(hotel_bookings$market_segment)
unique(hotel_bookings$distribution_channel)
unique(hotel_bookings$is_repeated_guest)
unique(hotel_bookings$reserved_room_type)
#unique(hotel_bookings$assigned_room_type)
unique(hotel_bookings$deposit_type)
unique(hotel_bookings$customer_type)
unique(hotel_bookings$reservation_status)
```


```{r}
my_data<-hotel_bookings
```

```{r}
my_data$is_canceled <-as.factor(my_data$is_canceled)
my_data$hotel<-as.factor(my_data$hotel)
my_data$meal<-as.factor(my_data$meal)
my_data$is_repeated_guest<-as.factor(my_data$is_repeated_guest)
my_data$reserved_room_type<-as.factor(my_data$reserved_room_type)
my_data$deposit_type<-as.factor(my_data$deposit_type)
my_data$customer_type<-as.factor(my_data$customer_type)
my_data$reservation_status<-as.factor((my_data$reservation_status))
str(my_data)
```
```{r}


```

## Including Plots


