---
title: "Final Project on Joe Biden's Ratings "
description: |
  Little steps towards Final Project 
author: "Isha Akshita Mahajan"
date: 08-19-2021
output: 
  distill::distill_article:
    self_contained: false
draft: TRUE
---

#Code Setup
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
library(tidyverse)
library(ggplot2)
library(dplyr)
library(haven)

#load stata package and get information on variable columns 
data<- read_stata("../../_data/stata.DTA")
#load csv file 
csv<-read_csv("../../_data/abc_poll_2021.csv") 
```


#Data Cleansing

```{r}
#Update csv with renamed columns, separate two races,
updated_csv <- csv %>% 
  rename(
  "education"= "ppeduc5",
  "age"="ppage",
  "education_category"="ppeducat",
  "gender"="ppgender",                              
  "race_and_ethnicity"="ppethm",                     
  "household_size"="pphhsize",
  "household_income"= "ppinc7",
  "marital_status"="ppmarit5",
  "msa_status"="ppmsacat",
  "region_of_residence"="ppreg4",
  "living_ownership_status"="pprent",
  "state"="ppstaten",
  "work_status"="PPWORKA",
  "employement_status"= "ppemploy",
  "covid_response"="Q1_a",
  "immigration"="Q1_b",
  "economic_recovery"="Q1_c",
  "afghanistan"="Q1_d",
  "gun_violence"= "Q1_e",
  "crime"="Q1_f",
 "coronavirus_concern"= "Q2",
 "one_or_more_vaccine_dose_recieved"="Q3",
 "campaign_promises"="Q4",
 "next_12_months_views"="Q5",
 "political_identification"="QPID") %>% 
  separate(race_and_ethnicity,sep = ", ",into= c("race_1", "race_2")) %>% 
```

#Survey Information

This data was taken from the ABC News/Ipsos poll conducted from July 23 to July 24, 2021, by Ipsos. The poll sampled a population of 527 participants who were 18 years or older. The demographics were drawn from the US Census Bureau’s 2019 Community Survey conducted in both English and Spanish. The margin of error for the total sample is expected to be +-/-5.0 percentage at the 95% confidence level. 

The survey asked variety of questions to the participants which spanned from Joe Biden's handling of covid_19 and economic recovery to immigration and the withdrawal of troops from Afghanistan. The poll aims to overall get an understanding of Joe Biden's rating after being in office for six months. 

The list of questions in includes as renamed in the final dataset for the analysis are:

1.covid_response = Do you approve or disapprove of the way Joe Biden is handling...the response to the coronavirus(COVID-19)?
2. immigration= Do you approve or disapprove of the way Joe Biden is handling...immigration and the situation at the United States-Mexico border?
3. economic_recovery = Do you approve or disapprove of the way Joe Biden is handling...the economic recovery?
4. afghanistan = Do you approve or disapprove of the way Joe Biden is handling...the withdrawal of United States troops from Afghanistan?
5. gun_violence = Do you approve or disapprove of the way Joe Biden is handling...gun violence?
6. crime= Do you approve or disapprove of the way Joe Biden is handling...crime?
7. coronavirus_concern = How concerned are you that you or someone you know will be infected with the coronavirus?...
8. one_or_more_vaccine_dose = Have you personally received at least one dose of a coronavirus vaccine, or not?
9. campaign_promises = How would you rate the job Joe Biden is doing keeping his campaign promises?....
10. next_12_month_views = Thinking about the next 12 months, would you say you feel optimistic or pessimistic about the way things are going in this country?

The current data were weighted to adjust for gender by age,race/ethnicity, education, Census region, metropolitan status, household income and party identification. The weights_pid column reflects those values. The data was downloaded using the Roper I-poll service and can be found here:  https://ropercenter.cornell.edu/ipoll/study/31118538/questions#d952e1cc-5dbd-45ea-9801-80bd8c5fdcdf


#Exploratory Analysis 

Q1. What are the political leanings of the survey participants  

```{r}
#Use the count function to count the number of times each person identified themselves politically into one of the categories
demographic_analysis<- csv %>% 
  count(QPID,"gender","marital_status")
```

Ans: 176 Participants identified as democrats, 152 as republicans, 168 and independents and 31 participants either skipped or chose not to identified. 


