---
title: "sathvik_hw5"
description: |
  Data Visualizations
author: sathvik_thogaru
date: 08-26-2021
output:
  distill::distill_article:
    self_contained: false
draft: true
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```


## Importing data
This data set contains a single file which compares various booking information between hotels.

```{r}
library(skimr)
library(lubridate)
library(tidyverse)
library(patchwork)
```

importing the data and reading the top 5 rows to understand the dataset

```{r}
hotel_bookings <- read_csv("../../_data/hotel_bookings.csv")
head(hotel_bookings)
```


skim() is used to for getting summary statistics about variables in dataframe,tibbles,datatablesand vectors. It is mostly used with grouped dataframes
(source: https://cran.r-project.org/web/packages/skimr/vignettes/skimr.html)


```{r}
skim(hotel_bookings)
```


looking for type of hotels booked by guests

```{r}
(hotels_info<-data.frame(table(hotel_bookings$hotel)))
```


from the above summary statistics we can see there are a total of 119390 rows and 32 columns in the hotel_bookings dataset. 13 character variables, 18 numeric variables, and 1 date variable. there are a total of 4 missing values in the children variable. for the analysis now i will be using hotel, market segment, stays_in_weekend_nights and stays_in_week_nights.

## Varaible Description

hotel variable: type of hotel booked\
market segment : Market segment designation. In categories, the term “TA” means “Travel Agents” and “TO” means “Tour Operators”\
stays_in_weekend_nights : guest stayed at the hotel in weekend nights\
stays_in_week_nights : guest stayed at the hotel in week nights\
I am using the select() from the dplyr package which comes with tidyverse package and the piping for selecting columns


```{r}
(df <- hotel_bookings %>% 
  select(hotel,stays_in_weekend_nights,stays_in_week_nights,market_segment))
```

Here I used the pie chart to represent the relative size of the type of hotels in the hotel booking dataset.
The piechart shows that there are more city hotels booked (79330) than the resort hotels(40060). 
I have chosen this to better visualize the values as a whole. It can also be shown using the heights of bar in barchart.


In this part I used unique function to know the different categories in market segment.

```{r}
unique(df$hotel)
unique(df$market_segment)
```

I split the dataset into resort hotel data and the city hotel data to work on them individually and then compare

```{r}
resort_hotel <- filter(df, hotel == "Resort Hotel") 
city_hotel <- filter(df, hotel =="City Hotel") 
```


bookings in different market segments



At city hotel

```{r}
(city_hotel_market_segment <- city_hotel %>% 
  count(market_segment))


a <- ggplot(city_hotel_market_segment, aes(market_segment,n))+
geom_histogram(fill = "indianred", stat = "identity") + 
labs(title = "City Market segment", y = "no of bookings") + 
  theme_classic()+
    theme(axis.text.x = element_text(angle = 90, hjust = 1))+
    theme(plot.title = element_text(size = 14, hjust = 0.5, face = "bold"))
```

 At resort hotel

```{r}
(resort_hotel_market_segment <- resort_hotel %>% 
  count(market_segment))


b <- ggplot(resort_hotel_market_segment, aes(market_segment,n))+
geom_histogram(fill = "indianred", stat = "identity") + 
labs(title = "Resort_Market segment",y = "no of bookings") + 
  theme_classic()+
  theme(axis.text.x = element_text(angle = 90, hjust = 1))+
theme(plot.title = element_text(size = 14, hjust = 0.5, face = "bold"))

a+b
```

I plotted the bar graph using ggplot between the market segment and the no of bookings made by each group of market segment.
In the chart it shows that many customers booked through online TA in both hotels. when compared there were more direct bookings in the resort hotel compared to city hotel.
there are no bookings made by Aviation in the resort hotel type.
As the items are many to compare I have used the barchart. I tried exploring other tpes og graphs but considered barchart as best to visualize this data for both resort and city hotel data and their respective share in market segment groups. I tried combining these graphs and show side by side bars of resort and city hotel to better visualize, but i am unable to do it with two dataframes. Used patchwork package to display them side by side but in different graphs. thanks to patchwork package for tyhe help..

In the next visualization I used the facet wrap to display the same above visualiztion between two hotels in different groups.


```{r}
(df %>% group_by(hotel,market_segment) %>% count(market_segment))
```

```{r}
(ggplot(df, aes(hotel)) + 
  geom_histogram(fill = "indianred", stat = "count") + 
  labs(title = "Market segment") + 
    theme(plot.title = element_text(size = 14, hjust = 0.5, face = "bold"))+
  facet_wrap(vars(market_segment), scales = "free_y"))
```


how many number of days do people stay in the hotel?

Resort hotel

```{r}

resort_hotel$total_nights <- resort_hotel$stays_in_weekend_nights + resort_hotel$stays_in_week_nights
city_hotel$total_nights <- city_hotel$stays_in_weekend_nights + city_hotel$stays_in_week_nights


  
resort_number_of_days_guests_stay_data <- resort_hotel %>% 
  group_by(total_nights) %>% 
  count()

resort_number_of_days_guests_stay_data$relative_no_of_bookings <- resort_number_of_days_guests_stay_data$n/sum(resort_number_of_days_guests_stay_data$n)*100

  
city_number_of_days_guests_stay_data <- city_hotel %>% 
  group_by(total_nights) %>% 
  count()

city_number_of_days_guests_stay_data$relative_no_of_bookings <- city_number_of_days_guests_stay_data$n/sum(city_number_of_days_guests_stay_data$n)*100

city_number_of_days_guests_stay_data
resort_number_of_days_guests_stay_data


```

```{r}
c <- plot1<-ggplot(resort_number_of_days_guests_stay_data, aes(total_nights,relative_no_of_bookings))+
   geom_bar(stat = "identity", fill="indianred")+
  labs(x = "no of nights stayed at the hotel", y = "realtive no of bookings", title = "length of stay at the resort hotel")+
  theme_classic()+
    theme(plot.title = element_text(size = 14, hjust = 0.5, face = "bold"))+
  xlim(c(0,30))


```




```{r}
d <- plot1<-ggplot(city_number_of_days_guests_stay_data, aes(total_nights,relative_no_of_bookings))+
   geom_bar(stat = "identity", fill="indianred")+
  labs(x = "no of nights stayed at the hotel", y = "relative no of bookings", title = "length of stay at the city hotel")+
  theme_classic()+
    theme(plot.title = element_text(size = 14, hjust = 0.5, face = "bold"))+
  xlim(c(0,20))

c+d
```

I used geom_bar to find the relationship between length of stay at each hotel and relative percentage of guests.In resort hotel, we can see that guests stay for mostly two to three nights but there is increase in resort hotel graph after 5 nights it shows that guests tend to stay at resort hotel probably if its a weekend and then move out. Here the resort hotel managemnet can work on the average daily rate if guests stay more than 5 days 
In city hotel guests stay mostly for 3-4 nights only. 