---
title: "ankithw3"
description: |
  Data Wrangling in Hotel Bookings dataset
author: Ankit Kumar
date: 08-19-2021
output:
  distill::distill_article:
    self_contained: false
draft: true
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```


loading relevant libraries

```{r}
library(tidyverse)
```


```{r}
# reading hotel bookings data and viewing it

library(readr)
hotel_bookings <- read_csv("../../_data/hotel_bookings.csv")
View(hotel_bookings)
```

inspecting the first few rows of the dataset
```{r}
hotel_bookings %>% head()
```
Overview of the dataset

```{r}

# Dimension of the dataset
hotel_bookings %>% dim()

# Names of the columns in the dataset
hotel_bookings %>% names()

# Summary of the dataset
hotel_bookings %>% summary()

# Type of data in each column
hotel_bookings %>% str()

```

Are Na values present in the dataset?

```{r}
hotel_bookings %>% is.na()%>% sum()

# there are 4 Na values in the data set
```

Identifying the positions of NA values in the dataset

```{r}
hotel_bookings %>% is.na() %>% which(arr.ind = TRUE)
# arr.ind is TRUE because we need the positions of NA values in the Dataset
```

Na values are present in the 11th column. So, examining the rows which have NA values

```{r}
hotel_bookings[c(40601,40668,40680,41161), ]

# The children column has NA values for the above rows
```

Removing the rows which have Na values
```{r}
hotel_bookings <- hotel_bookings %>% filter(!is.na(children))

# checking whether data set still has Na values
hotel_bookings %>% is.na()%>% sum()

```

Data Transformation

```{r}
library(plyr)
count(hotel_bookings, "hotel")

# The number of city hotels in the data set is very large relative to the number of resort hotels.
```

Segregating hotel data set into city hotels and resort hotels

```{r}

city_hotel <- hotel_bookings %>% filter(hotel_bookings$hotel == "City Hotel")

resort_hotel <- hotel_bookings %>% filter(hotel_bookings$hotel == "resort Hotel")

```


Exploring City Hotels further

```{r}
count(city_hotel, "market_segment")

city_hotel_by_mkt <- city_hotel %>% group_by(market_segment)

```

How many booking were made with deposits?

```{r}
city_hotel_by_mkt %>% count("deposit_type")

city_hotel_by_mkt %>% count("is_canceled")
```

```{r}

city_prices_monthly <-select(city_hotel, arrival_date_month, adr)


city_prices_monthly %>% summarise(mean_adr = mean(adr))
```


