---
title: "Homework 5 - Data Visualization"
description: |
  This dataset is named Active Duty Marital Status (ADMS). This dataset has been published by government of United States for public usage. It has count details of singles,   single parents, and married resources of the 4 branches of the defence department for the month of April 2010.
author: Abhinav Kumar
date: 08-24-2021
output:
  distill::distill_article:
    self_contained: false
draft: true
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


**Task:**

*You are again to make a distill blog post to the course blog. Your goal is to build on the previous assignments, and create a data visualization using the dataset you have been working with in R. Then, use R Markdown to draft a document that tracks the creation of the data visualization. In particular, comment on and discuss:*

*1. What the visualization demonstrates.*

*2. Why you chose the visualization approach you selected, and what alternative approaches you considered but decided not to pursue.*

*3. What you wished, if anything, you could have executed but found limited capability to do.*



```{r}
library(tidyverse)
library(readxl)
library(stringr)
```

*Invoking the requisite libraries here to wrangle and cleanse the data.*
```{r}

data <- ("C:/Users/Abhinavkumar.DESKTOP-JP9HSBL/Desktop/DACSS Summer/ActiveDuty_MaritalStatus.xls")

sheet_name <-  excel_sheets(data)

data_adms <-  map_df(sheet_name, ~ read_excel(data, skip = 8, sheet = .x, col_names = c("d", "Pay_Grade", "Single_NoKids_Male", "Single_NoKids_Female", "t", "Single_WithKids_Male", "Single_WithKids_Female", "t", "Married_JointService_Male", "Married_JointService_Female", "t", "Married_Civilian_Male", "Married_Civilian_Female", rep("t", 4))) %>%
           mutate("Defence_Wing" = .x))
  
```
*The dataset "data_adms" has 148 obervatiosn/rows and 18 columns/variables*
```{r}
dim(data_adms)
```
```{r}
str(data_adms)
```
*The original raw data was in ".xls" format with 5 sheets. After importaning the data using purrr::map_df function, all the sheets have been statcked on top of another* *starting from sheet 1 to sheet 5, converting it to a single data frame. This has introduced a column and few rows with "NA" values. *

*Feature Set Description:*

*1) Pay_Grade                    : This gives the information whether resource is Enlisted, Officer, or Warranted category reource across grades 1 to 10 ofthe                                                 defence forces.*
                      
*2) Single_NoKids_Male           : Charatcetr feature with details of male defence resource who is single and does not have children*

*3) Single_NoKids_Female         : Character feature with details of male defence resource who is single and does not have children*

*4) t...5                        : Charatcer feature giving the aggregate value of singles (male/female) with no children*

*5) Single_WithKids_Male         : Character feature with details of male defence resource who is single and has children*

*6) Single_WithKids_Female       : Character feature with details of female defence resource who is single and has children*

*7) t...8                        : Charatcer feature giving the aggregate value of singles (male/female) with children*

*8) Married_JointService_Male    : Character feature with details of male defence resource who is married to another defence resource in service*

*9) Married_JointService_Female  : Character feature with details of female defence resource who is married to another defence resource in service*

*10) t...11                      : Character feature giving the aggregate value of defence resources (male/female) married within defence force services*

*11) Married_Civilian_Male       : Character feature giving the details of male defence resource married to a civilian (not within defence forces)*

*12) Married_Civilian_female     : Character feature giving the details of female defence resource married to a civilian (not within defence forces)*

*13) t...14                      : Character feature giving the aggregate value of the defence resources married to civilians.*

*14) t...15                      : Aggregate of all the Males in the dataset across categories*

*15) t...16                      : Aggregate of all the Females in the dataset across categories*

*16) t...17                      : Aggregate of the males and females across categories*

*17) Defence_Wing                : Character feature giving outdetails to which arm of defence the resorce beloangs to (AirForce|MarineCorps|Navy|Army)*

*The first column "d" has got populated with "NA" values which reading the data from all the sheets of ".xls" document. It has no use for our analysis and will be dropped. Similarly the aggregate columns do not have any use and will be dropped. Within "Defence_Wing" *
```{r}

apply(data_adms, 2, function(x) any(is.na(x)))

```
*Here we are checking for NA values. All the features have NA values and will be dropped for the dataset.*
```{r}

data_adms <-  data_adms %>%
  select( "Pay_Grade" : "Single_NoKids_Female", "Single_WithKids_Male", "Single_WithKids_Female", "Married_JointService_Male", 
          "Married_JointService_Female", "Married_Civilian_Male", "Married_Civilian_Female",  "Defence_Wing") %>%
  na.omit() %>%
  filter(str_detect(Pay_Grade, "TOTAL", negate = T)) %>%
  filter(str_detect(Pay_Grade, "Pay Grade", negate = T)) %>%
  filter(str_detect(Defence_Wing, "TotalDoD", negate = T))%>%
  relocate(Defence_Wing) %>% 
  separate(col = Pay_Grade, into = c("Pay_Grade_Class","Pay_Grade"), sep="-") %>%
  mutate(Pay_Grade_Class = case_when(
    startsWith(Pay_Grade_Class, "E") ~ "Enlisted",
    startsWith(Pay_Grade_Class, "O") ~ "Officer",
    startsWith(Pay_Grade_Class, "W") ~ "Warrant")) %>%
  pivot_longer("Single_NoKids_Male" : "Married_Civilian_Female", names_to = "Status", values_to = "Count") %>%
  separate(col = Status, into = c("Marital_Status", "Family_Details", "Gender"))


```

```{r}
dim(data_adms)

str(data_adms)
```
*After cleansing the data, the dimension of the data frame is 720 observations/rows and 7 columns/features.*
*All the features are treated as characters by R. However, the "Count" feature is numeric. So we will coerce it to a numeric class.*
```{r}

data_adms$Count <- as.numeric(data_adms$Count)

str(data_adms)
```
*Finally all variables are in order. *

1) Defence_Wing                : Character feature giving out details to which arm of defence the resource belongs to (AirForce|MarineCorps|Navy|Army)

2) Pay_Grade_Class             : Categorical feature with 3 classes to give details whether the defence resource is Enlisted|Officer|Warranted

3) Pay_Grade                   : Categorical feature with 10 classes containing details of grades from 1 to 10

4) Marital_Status              : Categorical feature 2 classes (Single | Married)

5) Family_Details              : Categorical feature with 4 classes (NoKids | WithKids | JointService | Civilian)

6) Gender                      : Categorical feature with 2 classes (Female | Male)

7) Count                       : Numerical feature with count values of defence personnels under each category*
```{r}

df_temp <- as.data.frame(table(select(data_adms, "Defence_Wing", "Gender")))

df_temp

```
r
```{r}
data_adms %>% ggplot(aes(x = Defence_Wing, fill = Gender)) +
  geom_bar() + labs(title = "Gender Distribution across Defence_Wing") 

```
*Here I have plotted Defense_Wing versus the frequency of females and males across the 4 defense wings. I wanted to label the bars with the frequency value as well. Hence, I had prepared the data frame "df_temp" so that I can use the frequency column to label the stacked bar plot, but using "geom_text()" function was not very clear. So unable to use it.*


```{r}

data_adms %>% ggplot(aes(x = Defence_Wing, fill = Marital_Status)) +
                       geom_bar() + labs(title = "Marital Status Distribution across Defence_Wing") 
                         

```
*Here I have plotted Defense_Wing versus the frequency of Marital_Status across the 4 defense wings. The distribution of Married and Single defense personnels is same as distribution of Gender across all the 4 wings.*

```{r}
data_adms %>% ggplot(aes(x = Gender, fill = Family_Details)) +
                       geom_bar() + labs(title = "Family Details Distribution across Defence_Wing basis Gender") + facet_wrap(vars(Defence_Wing))
```
*Here I have plotted the distribution of Gender across across 4 wings of Defense using Family_Details to create stacked bar and Defense_Wing as the variable to create 4 plots for each wing of Defense.*
```{r}

data_adms %>% ggplot(aes(y = Defence_Wing, fill = Gender)) +
  geom_bar() + labs(title = "Pay_Grade_Class distribtion across Defence_Wing basis Gender") + facet_wrap(vars(Pay_Grade_Class))

```
*Here I have plotted the distribution of pay grade class across defense wings basis gender across the pay grade categories. Have used pay grade class as the variable to facet wrap.From the plot we can see that "Officer" category has the maximum defense personnels listed followed by "Enlisted". The distribution of gender across each category of pay garde class seems to be equal within the category.*

