---
title: "Zoe's Final Project Data"
description: |
 Introducing the final project data for homework 4 and 5
author: Zoe Bean
date: 08-20-2021
categories:
  - homework 4
  - homework 5
  - bicycle data
output:
  distill::distill_article:
    self_contained: false
draft: true
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Import datasets and load tidyverse package. This dataset is from the [Chicago Divvy Bicycle Sharing Data](https://www.kaggle.com/yingwurenjian/chicago-divvy-bicycle-sharing-data?select=data.csv) on Kaggle.

```{r import}
library(tidyverse)
bicycle_data<-read_csv("../../_data/chicago_bike_data.csv")
```

My questions: what type of customer is more likely to use the bike on bad weather days? what type of customer is more likely to have a longer trip? does weather impact how long the trips are? does a certain gender take longer trips? are there times of day one gender is more likely to bike?is a certain gender more likely to bike in different weather conditions?

## Homework 4

#### 1- Descriptions of the variables - how they were collected, any missing values, etc

First I Found out the possible columns in the dataset. Then find out the possible catagories for the events column,the gender column, and the usertype column. 

```{r analysis1}
colnames(bicycle_data)
unique(bicycle_data$events) 
unique(bicycle_data$gender) 
unique(bicycle_data$usertype) 
```


  'trip_id' is the unique id for each bike trip. 'year', 'month', 'day', and 'hour' are the date and time that each bike trip occured. 'week' is the week of the year each trip occurred at. This data was ranges from January 1, 2014 to January 1, 2017.
  
  'usertype' indicated what kind of customer is using the bike; 'Subscriber' is a customer that has purchased a yearly description, 'Customer' is a customer that has purchased this single ride, and I cannot find an indication on what type of customer is 'Dependant'. 
Gender is the gender of the person renting the bike, and it appears 'Male' and 'Female' were the only options for this dataset. 

  'starttime' is the date, hour, minute, and second the trip began, and 'stoptime' is the same information for when the trip ended. 'tripduration' is how long the trip lasted in minutes. 
  
  'tempurature' is the temperature at the time of the trip. 'events' is the weather event, if any, that occured during the trip: the options being "tstorms"(thunderstorms),"rain or snow","cloudy", "not clear", "clear", and "unknown".
  
  'from_station_id' is the id of the bike rental station at the beginning of the trip, and 'from_station_name' is the name of the station at that id.'to_station_id' is the id of the bike rental station at the end of the trip, and 'to_station_name' is the name of the station at that id.
  
  'latitude_start' and 'longitude_start' is the latitude and the longitude of the location where the trip began; 'latitude_end' and 'longitude_end' is the latitude and the longitude of the location where the trip ended.
  
  'dpcapacity_start' and 'dpcapacity_end' are the variables that indicate the charging dock capacity at the stations.
  
  All of the variables that are directly related to the bike trip were collected off of the [divvy bikeshare website](https://www.divvybikes.com/system-data). All of the weather related data was collected by the [weather underground website](https://www.wunderground.com/). The person that uploaded this dataset added only the relevant weather data and cleaned up some of the missing variables.
  
#### 2- How you cleaned and coded the data, including a before/after comparison as needed

I have no need for the following variables: longitudes, latitudes, and columns to do with stations, because I do not have any questions regarding the location or the specific bike stations. I also do not need the endtime, since we have start time and duration. 'week' is both redundant due to having the date in other variables, as well as being a confusing metric for time.

```{r remove cols}
bicycle_data_clean<-select(bicycle_data, 1:3, 5:9,11:13)
colnames(bicycle_data_clean)
```


I create a new column called 'weather_type' with the categories: extreme_temp, low_vis, safe, and dangerous. I also get rid of instances when the weather is 'unknown'.

I determine if the weather is an extreme temperature with the understanding that the average humidity of Chicago is 71 percent [source](https://www.chicagotribune.com/weather/ct-wea-asktom-0324-20180323-column.html). From there, I use the [heat index](https://www.weather.gov/ama/heatindex) and notice that at 70 percent humidity, the 'take extreme caution' catageory starts at 86 degrees Fahrenheit. So, it is extremely hot above 86 degrees.
I then calculate extreme cold with the understanding that the wind speed is 10mph [source](https://en.wikipedia.org/wiki/Climate_of_Chicago), (under 'windy city' it states that the average wind chill is from 8 to 12 mph, and I took the average of that). Then I look at this [Wind Chill Chart](https://www.weather.gov/safety/cold-wind-chill-chart)  and notice that at 10mph, frostbite in 10 minutes begins at -5 degrees Fahrenheit. Thus, extreme cold is below -5. So, an extreme temperature is when 'temperature' is 86 or above, or when it is -5 or below.

```{r extreme temp}
bicycle_data_clean<- bicycle_data_clean %>% mutate(extreme_temp=((temperature<=-5)|(temperature>=85)))
```

Now, I remove the instances where 'events' is 'unknown'. I consider it to be 'low_vis' then 'events' is anything but 'clear'.

```{r low vis}
bicycle_data_clean<-bicycle_data_clean %>% filter(events!='unknown') %>%
  mutate(low_vis=(events!='clear'))
head(bicycle_data_clean)
```

Now I take my 'low_vis' and 'extreme temp' catagories, and use them to make my weather_type category. If it is both 'low_vis' and 'extreme_temp', it is 'dangerous'. If it is neither, it is 'safe' and if it is just one it will be that catagory. I then remove my temporary variables as well as 'temperature' and 'events'.

```{r make column}
bicycle_data_clean<-bicycle_data_clean %>%
  mutate(weather_type = case_when(
         (low_vis == TRUE) & (extreme_temp==TRUE)  ~ "dangerous",
         (low_vis == FALSE) & (extreme_temp==FALSE)  ~ "safe",
         (low_vis == TRUE) & (extreme_temp==FALSE)  ~ "low_vis",
         (low_vis == FALSE) & (extreme_temp==TRUE)  ~ "extreme_temp"
         )) %>% select(1:9, 14)
  
head(bicycle_data_clean)
head(bicycle_data)
```

#### 3- summary descriptives of the recoded variables

Here, I look at the average trip duration and hour of the day for each weather type, gender, and usertype.

```{r summary}
bicycle_data_clean %>%
  group_by(weather_type) %>%
  select(tripduration, hour) %>%
  summarize_all(mean, na.rm = TRUE)

bicycle_data_clean %>%
  group_by(gender) %>%
  select(tripduration, hour) %>%
  summarize_all(mean, na.rm = TRUE)

bicycle_data_clean %>%
  group_by(usertype) %>%
  select(tripduration, hour) %>%
  summarize_all(mean, na.rm = TRUE)



```

## Homework 5


### Question: what type of customer is more likely to use the bike on bad weather days?

```{r userweather}
prop.table(xtabs(~ usertype + weather_type, bicycle_data_clean))
freq_table_user_weather <- bicycle_data_clean %>%
  select(usertype, weather_type) %>%
  table()
freq_table_user_weather
chisq.test(freq_table_user_weather)

```
#### 1- What the visualization demonstrates.

The proportion table as well as the frequency table show that there is a higher proportion of Subscribers riding in dangerous weather and low visibility, and the no Customer/Dependant is riding in extreme temperature. With a chi squared test showing a p-value well below .005, this difference is statistically significant.

#### 2- Why you chose the visualization approach you selected, and what alternative approaches you considered but decided not to pursue.

I chose a proportion table with a chi squared test because it answers my question and I could not find a better way to represent two categorical variables with possible influence on each other. 

#### 3- What you wished, if anything, you could have executed but found limited capability to do.

I wish I knew more ways to visualize and compare categorical data. 

### Question: is a certain gender more likely to bike in different weather conditions?

```{r genderweather}
prop.table(xtabs(~ gender + weather_type, bicycle_data_clean))
freq_table_gender_weather <- bicycle_data_clean %>%
  select(gender, weather_type) %>%
  table()
freq_table_gender_weather
chisq.test(freq_table_gender_weather)
```
#### 1- What the visualization demonstrates.

The proportion table shows there is a higher proportion of men riding in dangerous weather and low visibility. With a chi squared test showing a p-value well below .005, this difference is statistically significant. In fact, the proportioin table shows that men are about twice as likely to go out in dangerous weather and around three times as likely to go out in extreme temperatures.

#### 2- Why you chose the visualization approach you selected, and what alternative approaches you considered but decided not to pursue.

I chose a proportion table with a chi squared test because it answers my question and I could not find a better way to represent two categorical variables with possible influence on each other. 

#### 3- What you wished, if anything, you could have executed but found limited capability to do.

I wish I knew more ways to visualize and compare categorical data.
### Question: are there times of day one gender is more likely to bike?

```{r genderhour}
ggplot(bicycle_data_clean, aes(hour, fill=gender)) +
  geom_bar()+labs(title = "Gender by Hour", y="gender count")
ggplot(bicycle_data_clean, aes(hour, fill=gender)) +
  geom_bar(position="fill")+labs(title = "Gender Frequency by Hour", y="gender frequency")
```
#### 1- What the visualization demonstrates.

The first visualization shows the count of people who bike at a certain hours of the day, color stacked by gender. This is bimodal, showing people biking most frquently at around 7/8am, and 4/5pm(or 16 and 17). These peaks likely are due to people going to and returning from work or school. Overall, very few people bike before 5am. 

The second visualization shows the proportion of each gender at each hour.  On average, there are about 80% men biking at any hour of the day. Interestingly, at 5am, there is a sudden increase in the proportion of women biking, although it does not appear that it is statistically significant.

#### 2- Why you chose the visualization approach you selected, and what alternative approaches you considered but decided not to pursue.

I chose two visualizations to show this because I wanted to show both the overall trend of what hour of the day is popular AND when the proportion of each gender fluctuates. I had to do two because while ideally, the first visualization would show what I want, there are so much more men than women and the count of people per hour changes so drastically that you really can't see how the proportion fluctuates properly.

#### 3- What you wished, if anything, you could have executed but found limited capability to do.

I really wish that I could have shown the overall trends of the biking hourly as well as ny fluctuations in the proportions of gender at the same time, but the vast gap between the amount of men riding and women riding made that difficult to do.

### Question: what type of customer is more likely to have a longer trip?

```{r userduration}
bicycle_data_clean %>%
  group_by(usertype) %>%
  select(tripduration) %>%
  summarize(mean(tripduration), na.rm = TRUE) %>%
  ggplot(aes(x=`mean(tripduration)`, y=usertype, fill=usertype)) +
  geom_bar(stat="identity")+labs(title = "Users by Mean Trip Duration", x=" mean trip duration in minutes")
```
#### 1- What the visualization demonstrates.

This chart show how long, on average, each type of customer rides the bike for. The one time customer tends to have the longest trips, at a little over 15 minutes, and the year-long subscriber has the shortest trips on average, at around 11 minutes. This makes sense, as the subscriber would likely take more frequent, shorter trips to make the most of the subscriptions, while the one-time customer would either be unfamiliar with the route(thus taking longer), or would want to make the most of a single trip since that is what they are paying for.

#### 2- Why you chose the visualization approach you selected, and what alternative approaches you considered but decided not to pursue.

I chose to show this bar graph of the mean times because that way you can see how the customer type could influence the length of the ride. If I did not have it as a mean, you would only be seeing information about the subscribers since they are by far more common than the others.

#### 3- What you wished, if anything, you could have executed but found limited capability to do.

I think this viaualization really conveys what I want it to. The only thing I would like is to know what 'dependant' means so that I can interperet that catagory as well. I also would like to have more data points in the other catagories for more accuracy.

### Question: does a certain gender take longer trips?

```{r genderduration2}
ggplot(bicycle_data_clean , aes(x=tripduration, fill=gender)) + geom_density(alpha=0.4)+labs(title = "Gender by Trip Duration", x="trip duration")
```

#### 1- What the visualization demonstrates.

This density chart shows that while most trips are around 5 minutes long, a higher proportion of men are taking these five minute treks. Women are ever-so-slightly more likely to take trips between 10 and 30 minute trips than men.

#### 2- Why you chose the visualization approach you selected, and what alternative approaches you considered but decided not to pursue.

Originally, I was going to do a box plot to show the distributions for men and women, but as you can tell from the example below, the differences were so subtle that they are hard to pick out in that format, as well as the long tail manifesting itself in a distracting way in this format. Because of how male-dominated this dataset is, a bar graph does not really work.


```{r genderduration1}
bicycle_data_clean %>%
  group_by(gender) %>%
  select(tripduration) %>%
  ggplot(aes(y=tripduration, x=gender, fill=gender)) +
  geom_boxplot()+labs(title = "Gender by Trip Duration", x=" mean trip duration")
```

#### 3- What you wished, if anything, you could have executed but found limited capability to do.

I do think this visualization answers my question and shows the results in the best way I could show them. 


### Question: does weather impact how long the trips are?

```{r weatherduration}
bicycle_data_clean %>%
  group_by(weather_type) %>%
  select(tripduration) %>%
  ggplot(aes(x=tripduration, y=weather_type, fill=weather_type)) +
  geom_violin()+labs(title = "Trip Duration by Weather", x="trip duration")

```

#### 1- What the visualization demonstrates.

Overall, I think this visualization shows that no, weather does not really influence the length of a trip. All of the violins are similar in shape and have similar points of interest. The one with the most difference is not the 'dangerous' category, but the 'extreme_temp' category, with slightly less short trips and a little more long trips, but even that is negligible.

#### 2- Why you chose the visualization approach you selected, and what alternative approaches you considered but decided not to pursue.

I chose this approach because it shows a clearer distribution than a box plot or bar graph. The density plot ended up so similar it did not end up visually conveying anything about the weather categories.  

#### 3- What you wished, if anything, you could have executed but found limited capability to do.

Since there is little difference between the categories, I could really use most visualizations if I wished to, so there is nothing I really need to do- there is no 'ideal visualization' that I would be reaching for, so there is nothing more I wish to do with it.

