---
title: "Marina's HW5"
description: |
  Visualization homework with ggplot2.
author:
  - name: Marina
date: 08-25-2021
output:
  distill::distill_article:
    self_contained: false
draft: true
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

The requirement for HW5 is to build on the previous homework and develop a relevant visualization.

One of the last things I did for HW4 was to flatten part of the Lifeline dataset, accounting for adjustments made over time to claimed subscriber amounts. This left me with a dataset breaking down total claimed subscribers by each company, month by month, since 2018.

```{r load libraries and bring in data, echo=FALSE}
  library(RSocrata)
  library(tidyverse)

  df <- read.socrata("https://opendata.usac.org/resource/tfxa-gt3x.json")
  df <- filter(df, state == "MA" & support_year >= 2018)

```

```{r flatten, echo=FALSE, warning=FALSE, message=FALSE}
#Rename companies, but only if the total number of companies is 6
  #(otherwise there might have been name changes that I'd have to
  #account for)
  if (length(unique(df$sac_name)) == 6) {
  df <- df %>%
        mutate(sac_name = str_to_title(sac_name)) %>%
        mutate(provider = case_when(
              startsWith(sac_name, "City") ~ 'Westfield',
              startsWith(sac_name, "Global") ~ 'StandUp Wireless/Global Connection',
              startsWith(sac_name, "Granby") ~ 'Granby',
              startsWith(sac_name, "Tracfone") ~ 'Tracfone/Safelink',
              startsWith(sac_name, "Verizon") ~ 'Verizon',
              startsWith(sac_name, "Virgin") ~ 'Assurance/Virgin/T-Mobile'
              ))
    }  else {
      print("There has been a change in the number of unique companies")
    }
  subscribers <- df %>%
                  filter(!submission_type == "ADJUSTMENT") %>% #Exclude fin. adj. rows
                  select(provider, 
                         support_year,
                         support_month = support_month_date_format,
                         technology_type,
                         service_type,
                         subscriber_count) %>%
                  group_by(provider,
                           support_year,
                           support_month,
                           technology_type,
                           service_type) %>%
                  summarise(subscriber_count = sum(as.numeric(subscriber_count)))
```
```{r quick peek}
head(subscribers)
```

We'll start by visualizing overall totals across Massachusetts, without minding providers or types of service.

```{r Graph 1: Overall Total, echo=TRUE}
#Proper capitalization
  subscribers <- subscribers %>%
                  mutate(technology_type = str_to_sentence(technology_type),
                         service_type = str_to_sentence(service_type))

#Identify where data is missing
  missing_check <-  subscribers %>%
                    ungroup() %>% #If I don't do this, it'll add needless variables
                    select(provider, support_month) %>%
                    distinct() %>%
                    select(support_month) %>%
                    count(support_month) %>%
                    arrange(desc(support_month))
  
  #Identify latest month of complete data
    
    for (i in 1:5) { #check 5 latest months)
      if (missing_check$n[i] == 6) {
        full_month <- missing_check$support_month[i]
        break
      }
    }
  
  #Graph overall totals, excluding the recent months that aren't complete
  waiver_end_date <-  as.POSIXct("2021-05-01") #date that waiver expired 
  p1 <- subscribers %>%
            filter(support_month <= full_month) %>% #keep data up to most recent complete month
            group_by(support_month) %>%
            summarise(subscribers = sum(subscriber_count)) %>%
          ggplot(aes(x=support_month, y=subscribers)) +
            geom_line() +
            xlab("Support Month") +
            scale_y_continuous(name="Number of Subscribers", limits=c(0, 200000)) +
            geom_vline(xintercept = waiver_end_date, linetype="dotted", 
                color = "red", size=1.5) +
            annotate("text", x = (waiver_end_date - 50), y = 155000, 
                     label = "End of Waiver \n", angle=90) +
            ggtitle("Overall Amount of Lifeline Subscribers")
  
  p1 + theme_classic()
  
```
I added a line to show the end of a federal covid-related waiver, which was preventing companies from removing subscribers who hadn't been actively using the service in the past couple of months. The end of the waiver is likely a big factor in why we see a sharp drop in subscribers.

Next, let's look at totals by types of service offered, to see if there are any differences in enrollment trends.

```{r Graph 2: service type totals, echo=TRUE}
 p2 <- subscribers %>%
              ungroup() %>%
              select(support_month, service_type, subscriber_count) %>%
              filter(support_month <= full_month) %>% 
              group_by(support_month, service_type) %>%
              summarise(subscribers = sum(subscriber_count)) %>%
        ggplot(aes(x=support_month, y=subscribers,
                   group = service_type, color = service_type)) +
            geom_line() +
            xlab("Support Month") +
            scale_y_continuous(name="Number of Subscribers", limits=c(0, 150000)) +
            geom_vline(xintercept = waiver_end_date, linetype="dotted", 
                color = "red", size=1.5) +
            labs(col="Service Type") +
            theme_classic() +
            annotate("text", x = (waiver_end_date - 50), y = 130000, 
                     label = "End of Waiver \n", angle=90) +
            ggtitle("Lifeline Subscribers by Service Type")

p2
```

Although all service types have experienced important drops, it looks like voice service has remained the most stable throughout 2020 and 2021. 

### Reflection
* It's always a challenge to decide what kind of chart to make, and how many different things to try to plot at once. I could have tried using `facet_wrap` to divide the graphs between different service types, but I was curious to see how they all compared against each other, so decided to keep them all in one place.
* I'm not happy with the 'end of waiver' annotation I added to the red line, because I ended up manually choosing the positioning that made the most sense. I wonder if there isn't an easier/more automated way to choose an annotation position relative to `geom_vline`.
* I'm going to work on creating additional charts with different cuts of data. It's a bit of a chicken and egg problem, though, in a way-- sometimes it's hard to imagine how you want to cut the data without seeing a chart first, and sometimes it's not until you see a specific chart that you get an idea for a different cut.